---
title: "NCR RNAseq - enrichment analysis of gene lists of interest"
author: "WD, NW"
description: "This R code is to replicate the analyses, figures and tables from the research article entitled: Multi-insecticide resistant malaria vectors in the field remain susceptible to malathion, despite the presence of Ace1 point mutations, 2021"
date: "9/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Importing libraries :
```{r import }
library(fgsea)
library(data.table)
library(plotly)
library(tidyverse)

# set baseFolder / working directory to a "data" folder where all relevant STables "input_data" are saved as .txt files.
# baseFolder='~/data/'

```

## 1. Loading up the data

```{r load_gene_list}

geneLists = read.table( paste( baseFolder , 'NCR_annotation_V4_gsea_new4real.txt' , sep = '/' ) , header=T ) 
# 93 P450 clan = 90 P450 + 2 x cyt b5 + NADPH red 

geneLists <- geneLists%>%
  rename(enzyme_family = new_gsea)

head(geneLists)

table( geneLists$enzyme_family )
```

```{r load_DE_file}
## change comments depending on the comparison one wants to look at

# fileName = 'contrast_Field_Lab_C.edgeR_groupedModel.allGenes.txt' #comparison of 3 field populations grouped versus 2 lab grouped

# fileName = 'contrast_Agb_Lab_C.edgeR_mixedModel.allGenes.txt' #comparison of Agboville (NEC) versus 2 lab grouped
# fileName = 'contrast_Dab_Lab_C.edgeR_mixedModel.allGenes.txt' #comparison of Dabou (NEC) versus 2 lab grouped
fileName = 'contrast_Tia_Lab_C.edgeR_mixedModel.allGenes.txt' #comparison of Tiassalé (NEC) versus 2 lab grouped

DEinfo = read.table( paste( baseFolder , '03_DGE_analysis_results' , fileName , sep='/' ) , header=T )

summary(DEinfo)
```

Declaring a gene DE if the fdr is 0.01 
and with no LFC threshold.
```{r}

DEgenes = rownames( DEinfo )[ DEinfo$FDR < 0.01 ]
print( paste( 'number of DE genes :' , length(DEgenes) ))
```
...



## 2. gsea curve

Using the [fgsea package](http://bioconductor.org/packages/release/bioc/html/fgsea.html)

Performing a GSEA analysis :
```{r}

gene_lists = list()

for( i in unique(geneLists$enzyme_family) )
{
  if( !is.na(i) ){
  if( sum(geneLists$enzyme_family == i, na.rm = TRUE) >10 ){
  gene_lists[[ i ]] = geneLists$gene_id[ geneLists$enzyme_family == i ]    
  }}
}

gene_lists[['cuticular_protein_CPR']]
DEinfo[  gene_lists$cuticular_protein_CPR , ]

DEinfo[  gene_lists[['cuticular_protein_CPR']]  , ]
sum ( !is.na( DEinfo[ gene_lists$P450, 'logFC' ]  ) )
sum ( !is.na( DEinfo[ gene_lists$cuticular_protein_CPR, 'logFC' ]  ) )


lfc = DEinfo$logFC
ranks = lfc
names(ranks) = rownames(DEinfo)

gseaRes = fgsea( gene_lists , ranks , nperm=10000 )
print(gseaRes)

```

Lists COE and P450 seems to be significantly enriched.


I also perform a fisher test, to complement the GSEA whose metric is not ideal (here, it only accounts for log2FDR (should be log2FC), not for FDR) :
```{r}


nbSets = length(gene_lists)
fisherTestEnrichment = data.frame( nbGenes = rep(0,nbSets)   ,
                                   row.names = names(gene_lists) )
for ( i in names( gene_lists ) )
{
 fisherTestEnrichment[ i , 'nbGenes' ] = length(gene_lists[[i]])
}

PV = c()
DE = c()
for ( i in names( gene_lists ) )
{
  general = c(  sum( DEinfo$FDR >= 0.01 ) , sum( DEinfo$FDR < 0.01 ) ) 
  set = c(  sum( DEinfo[ gene_lists[[i]] , 'FDR']  >= 0.01  , na.rm=T)  , sum( DEinfo[ gene_lists[[i]] , 'FDR']  < 0.01  , na.rm=T) )
    
  r = fisher.test( matrix( c(general , set)  , nrow=2 , byrow=T) )
  DE = c( DE , set[2] )
  PV = c( PV , r$p.value )
    
}
fisherTestEnrichment['nbDE'] = DE
fisherTestEnrichment['pval'] = PV
fisherTestEnrichment['FDR'] = p.adjust(PV , method = 'BH')


print('general population :')
print('non-DE DE')
print(general)

fisherTestEnrichment
```
When adjusting for multiple testing, no set is seen as significantly enriched, given that a large fraction of genes are significantly DE.

However, note that some small sets have a all their genes enriched, so further enquiry may still be of interest.



```{r echo=FALSE}


makeInteractiveGSEAplot = function( DEGgeneTable , geneVector , titleLabel, DE_threshold=0.01 , LFC_threshold=0.0  )
{
  ## defining the statistic used for ranking in gsea
  stats = DEGgeneTable$logFC
  names(stats) = rownames(DEGgeneTable)
  pathway = geneVector
  
  ## getting the ES curve from fgsea
  rnk <- rank(-stats)
  ord <- order(rnk)
  statsAdj <- stats[ord]
  statsAdj <- sign(statsAdj) * (abs(statsAdj)^1)
  statsAdj <- statsAdj/max(abs(statsAdj))
  pathway <- unname(as.vector(na.omit(match(pathway, names(statsAdj)))))
  pathway <- sort(pathway)
  gseaRes <- calcGseaStat(statsAdj, selectedStats = pathway, 
                          returnAllExtremes = TRUE)
  tops = gseaRes[['tops']]
  bottoms = gseaRes[['bottoms']]
  
  ## defining log2FC breaks projected onto ranked log2FC
  posBreaks = c()
  lblBreaks = 5:-5
  for ( i in lblBreaks )
  {
    posBreaks  = c(posBreaks , Position(function(x) x <= i, stats[ord])  )
  }
  
  # defining DE statistical significance 
  statsSig =  as.factor( ( DEGgeneTable$FDR <= DE_threshold ) * sign(DEGgeneTable$logFC) )
    
  # ordering the gene in the geneSet of interest according to their LFC
  M = names( stats[ord] )  %in% geneVector
  orderedGeneSet = names(stats[ord][pathway])#names(stats[ord][M]) # ordered data in the set
  
  ## creating objects to plot the curve
  n <- length(statsAdj)
  xs <- as.vector(rbind(pathway - 1, pathway))
  ys <- as.vector(rbind(bottoms, tops))
  toPlot <- data.frame(rnk = c(0, xs, n + 1), es = c(0, ys, 0) )
  diff <- (max(tops) - min(bottoms))/8
  x = y = NULL
  
  ## plotly only
  pp <- plot_ly(data=toPlot, x=~rnk, y= ~es , width = 1000, height = 500)  %>%  
    layout(title = titleLabel, 
           xaxis = list( tickvals=posBreaks , ticktext=lblBreaks , title= "log2FC (ranked)"  ) , 
           yaxis=list(title='Enrichment Score')  ) %>% 
    add_segments( 
      x = (1:length(statsSig)), 
      y = -diff/4, 
      xend = (1:length(statsSig)), 
      yend = diff/4 , 
      color= c('underexpressed','non-significant','overexpressed')[statsSig[ord] ] , 
      colors = c('gray','red','blue') ,
      line=list(width = 0.25) , 
      hoverinfo="none") %>%
    add_lines( showlegend = FALSE, color = I("green"), hoverinfo = "none") %>%  
    add_segments(x = pathway, 
                 y = rep(-diff/2,length(pathway)), 
                 xend = pathway, 
                 yend = rep(diff/2,length(pathway)), 
                 color = I("black"), 
                 text = paste( paste( orderedGeneSet , DEGgeneTable[orderedGeneSet , 'gene_name'] ,sep=':' ) , 
                               paste('log2FC' , format(stats[orderedGeneSet] , digits=3) ),  
                               paste('FDR : ' , format(DEGgeneTable[ orderedGeneSet , 'FDR'] , digits=3 ) ) , 
                               sep='\n') ,
                 name = "genes in group"
                 )
  pp
}

```

## Visualisation of GSEA enrichement

I do this interactive visualization for all families with more than 10 genes :

```{r}
group="P450_clan"
pp = makeInteractiveGSEAplot( DEinfo , gene_lists[[group]] , titleLabel = paste("Field vs. Lab . Control -" , group ) ,  DE_threshold=0.01 , LFC_threshold=0.0  )
pp
```

```{r}
group="COE"
pp = makeInteractiveGSEAplot( DEinfo , gene_lists[[group]] , titleLabel = paste("Field vs. Lab . Control -" , group ) ,  DE_threshold=0.01 , LFC_threshold=0.0  )
pp
```

```{r}
group="GST"
pp = makeInteractiveGSEAplot( DEinfo , gene_lists[[group]] , titleLabel = paste("Field vs. Lab . Control -" , group ) ,  DE_threshold=0.01 , LFC_threshold=0.0  )
pp
```

```{r}
group="UGT"
pp = makeInteractiveGSEAplot( DEinfo , gene_lists[[group]] , titleLabel = paste("Field vs. Lab . Control -" , group ) ,  DE_threshold=0.01 , LFC_threshold=0.0  )
pp
```


```{r}
group="ABC_transporter"
pp = makeInteractiveGSEAplot( DEinfo , gene_lists[[group]] , titleLabel = paste("Field vs. Lab . Control -" , group ) ,  DE_threshold=0.01 , LFC_threshold=0.0  )
pp
```


```{r}
group="ribosomal_protein"
pp = makeInteractiveGSEAplot( DEinfo , gene_lists[[group]] , titleLabel = paste("Field vs. Lab . Control -" , group ) ,  DE_threshold=0.01 , LFC_threshold=0.0  )
pp
```

```{r}
group="cuticular_protein_CPLC"
pp = makeInteractiveGSEAplot( DEinfo , gene_lists[[group]] , titleLabel = paste("Field vs. Lab . Control -" , group ) ,  DE_threshold=0.01 , LFC_threshold=0.0  )
pp
```

```{r}
group="H+_transporting_ATP_synthase"
pp = makeInteractiveGSEAplot( DEinfo , gene_lists[[group]] , titleLabel = paste("Field vs. Lab . Control -" , group ) ,  DE_threshold=0.01 , LFC_threshold=0.0  )
pp
```

